\documentclass[a4paper,11pt]{article}
\usepackage[scale={0.8,0.9},centering,includeheadfoot]{geometry}
\usepackage{amstext}
\usepackage{listings}
\usepackage{verbatim}
\begin{document}

\section*{Autoregressive model of order $p$ (AR$(p)$)}

\subsection*{Parametrization}

The autoregressive model of order $p$ (AR1$(p)$) for the Gaussian vector
$\mathbf{x}=(x_1,\dots,x_n)$ is defined as (in obvious notation)
\begin{displaymath}
    x_{t} = \phi_{1}x_{t-1} + \phi_{2}x_{t-2} + \cdots + \phi_{p}
    x_{t-p} + \epsilon_{t}
\end{displaymath}
for $t = p, \ldots, n$, and where the innovation process
$\{\epsilon_{t}\}$ has fixed precision.

The AR$(p)$ process has an awkward parameterisation, as there are
severe non-linear constraints on the $\phi_{\cdot}$-parameters for it
to define a stationary model. Therefore we re-parameterized using the
partial autocorrelation autocorrelation function,
$\{\psi_{k}, k=1, \ldots, p\}$, where $|\psi_{k}|< 1$ for all $k$
\footnote{See for example
    \texttt{https://en.wikipedia.org/wiki/Partial\_autocorrelation\_function}.
    For $p=1$, then $\psi_{1} = \phi_{1}$, and for $p=2$, then
    $\psi_{1} = \phi_{1}/(1-\phi_{2})$ and $\psi_{2} = \phi_{2}$.} and
its \emph{marginal (NOT conditional) precision} $\tau$. Furthermore,
the joint distribution for $\{x_{t}, t=1, \ldots, p\}$, is set to the
stationary distribution for the process, hence there are no boundary
issues.

\subsection*{Hyperparameters}

The marginal precision parameter $\tau$ is represented as
\begin{displaymath}
    \theta_1 =\log(\tau) 
\end{displaymath}
and the prior for the marginal precision is defined on
$\theta_{1}$. The partial autocorrelation function $\{\psi_{k}\}$ is
represented
\begin{displaymath}
    \psi_{k} = 2\frac{\exp(\theta_{k+1})}{1+\exp(\theta_{k+1})} -1
\end{displaymath}
for $k = 1, \ldots, p$. The prior for $\{\theta_{k+1}, k=1, \ldots,
p\}$ is \emph{defined} to be multivariate normal with mean $\mu$ and
precision matrix $Q$.

\subsection*{Specification}

The AR$(p)$ model is specified inside the {\tt f()} function as
\begin{verbatim}
 f(<whatever>, model="ar", order=<p>, hyper = <hyper>)
\end{verbatim}
The option \texttt{order} ($>0$) is required. The multivariate normal
prior for $\{\theta_{k+1}, k=1, \ldots, p\}$, \texttt{is specified} as
the parameters to the prior for $\theta_{2}$ (the first
pacf-parameter), and the parameters to the multivariate normal prior
(\texttt{mvnorm}), is \texttt{c($\mu, Q$)}; see the example below.

\subsubsection*{Hyperparameter spesification and default values}
\input{../hyper/latent/ar.tex}

\clearpage
\subsection*{Example I}
\verbatiminput{example-ar.R}

\clearpage
\subsection*{Example II}

I this example we demonstrate how the parameters in the PC-prior for
AR$(p)$ can be set using the idea in {S.~H.~S{\o}erbye and H.~Rue
    (2016)}\footnote{Penalised complexity priors for stationary
    autoregressive processes. arXiv preprint arXiv:1608.08941, UiT The
    Artic University of Norway, 2016}.
    
\verbatiminput{ar-demo.R}

\clearpage
\subsection*{Notes}

\begin{itemize}
\item The functions \texttt{inla.ar.pacf2phi} and
    \texttt{inla.ar.phi2pacf} converts from the the $\phi$-parameters
    to the $\psi$-parameters, using the Durbin-Levinson
    recursions. These can also be used to compute, the marginal
    posteriors of the $\phi$-parameters from an approximation of the
    joint of the $\phi$-parameters; see the example for a simulation
    based approach.
\item Currently, the order $p$ is limited to $10$. If this creates a
    problem, let us know.
\item If some of the $\psi_{k}$-parameters are fixed, and $k < p$,
    then the marginal (log-)likelihood is wrong; The joint normal
    prior for all the $p$ $\psi$-parameters is used and not the
    conditional normal prior condition on the fixed
    $\psi_{k}$-parameters. If this creates a problem, let us know.
\item The prior spesification for the multivariate normal is a bit
    awkward. Hopefully, we will come up with a better way to do this
    in the future.
\end{itemize}

\end{document}

% LocalWords:  Autoregressive Parametrization autoregressive parameterisation

%%% Local Variables: 
%%% TeX-master: t
%%% End: 
% LocalWords:  parameterized autocorrelation Hyperparameters mvnorm ar
% LocalWords:  Hyperparameter spesification Durbin Levinson
