\documentclass[a4paper,11pt]{article}
\usepackage[scale={0.8,0.9},centering,includeheadfoot]{geometry}
\usepackage{amstext}
\usepackage{listings}
\begin{document}

\section*{Random walk model of order $1$ (RW1)}

\subsection*{Parametrization}

The random walk model of order $1$ (RW1) for the Gaussian vector
$\mathbf{x}=(x_1,\dots,x_n)$ is constructed assuming independent
increments:
\[
\Delta x_i = x_i-x_{i-1}\sim\mathcal{N}(0,\tau^{-1})
\]
The density for $\mathbf{x}$ is derived from its $n-1$ increments as
\begin{eqnarray*}
    \pi(\mathbf{x}|\tau) &\propto& \tau^{(n-1)/2}
    \exp\left\{-\frac{\tau}{2}\sum (\Delta x_i)^2\right\}\\
    & = &\tau^{(n-1)/2}\exp\left\{-\frac{1}{2}
      \mathbf{x}^T\mathbf{Q}\mathbf{x} \right\}
\end{eqnarray*}
where $\mathbf{Q}=\tau\mathbf{R}$ and $\mathbf{R}$ is the structure
matrix reflecting the neighbourhood structure of the model.

It is also possible to define a {\it cyclic} version of the RW1 model,
in this case the graph is modified so that last node $x_n$ is
neighbour of $x_{n-1}$ and $x_1$.
\subsection*{Hyperparameters}

The precision parameter $\tau$ is represented as
\begin{displaymath}
    \theta =\log \tau
\end{displaymath}
and the prior is defined on $\mathbf{\theta}$.

\subsection*{Specification}

The RW1 model is specified inside the {\tt f()} function as
\begin{verbatim}
 f(<whatever>, model="rw1", values=<values>, cyclic=<TRUE|FALSE>,
              hyper = <hyper>, scale.model = FALSE)
\end{verbatim}
The (optional) argument {\tt values } is a numeric or factor vector
giving the values assumed by the covariate for which we want the
effect to be estimated. See next example for an application.
 
The logical option \verb|scale.model| determine if the model should be
scaled to have an average variance (the diagonal of the generalized
inverse) equal to 1. This makes prior spesification much
easier. Default is \verb|FALSE| so that the model is not scaled.

\subsubsection*{Current recommended prior}
If you do not know which prior to use, the current recommendation is
\begin{verbatim}
 u = 1
 f(<whatever>, model="rw1", scale.model = TRUE
               hyper = list(theta = list(prior="pc.prec", param=c(u,0.01))))
 inla.doc("pc.prec")
\end{verbatim}
where {\tt u} should be set to a value appropriate for your case:
\begin{description}
\item[Gaussian likelihood (no link)] Set {\tt u} to be the empirical standard deviation of your data
\item[Poisson likelihood and log link] Set {\tt u} to 1
\item[Binomial and logit link] Set {\tt u} to 0.5
\item[Binomial and probit link] Set {\tt u} to 0.33
\end{description}
Increasing {\tt u} gives a weaker prior, decreasing {\tt u} gives a stronger prior.

\subsubsection*{Hyperparameter spesification and default values}
\input{../hyper/latent/rw1.tex}


\subsection*{Example}

\begin{verbatim}
n=100
z=seq(0,6,length.out=n)
y=sin(z)+rnorm(n,mean=0,sd=0.5)
data=data.frame(y=y,z=z)

formula=y~f(z,model="rw1",
            hyper = list(prec = list(prior="loggamma",param=c(1,0.01))))
result=inla(formula,data=data,family="gaussian")
\end{verbatim}


\subsection*{Notes}

\begin{itemize}
\item The RW1 is intrinsic with rank deficiency 1.
\item The RW1 model for irregular locations are supported although not
    described here.
\item The term $\frac{1}{2}\log(|R|^{*})$ of the normalisation
    constant is not computed, hence you need to add this part to the
    log marginal likelihood estimate, if you need it.
\end{itemize}
\end{document}


% LocalWords: 

%%% Local Variables: 
%%% TeX-master: t
%%% End: 
