---
title: "Installation of INLA and inlabru with Apptainer on HPC"
author: Man Ho Suen (m.h.suen@sms.ed.ac.uk)
output:
  rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Apptainer}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

(This vignette is a copy of the vignette in the *inlabru*-package.)

High Performance Computing (HPC) is surging in popularity. Building
INLA package on the HPC systems with the Anaconda environment
sometimes can be tricky due to the Linux distributions and glibc
versions. This article is to suggest a workaround with container
images using [Apptainer](https://apptainer.org/) (formerly known as
Singularity) because [Docker](https://www.docker.com/) requires root
access and is usually not available on HPC. Apptainer can pull Docker
images. Despite its convenience, the image size can be large.

## Installation Procedures 

1. Check if [Apptainer](https://apptainer.org/docs/user/main/cli/apptainer_version.html)
   is available on your HPC. If not, ask the administrator to install it. Type 
   on your HPC terminal to check which version is installed. 
```{r apptainer check, eval=FALSE}
$ apptainer version
```

2. Type the code below to pull the
[rocker/geospatial](https://rocker-project.org/images/versioned/rstudio.html)
Singularity Image Format (SIF) file with the latest R version. For
illustration, we'll assume the file is stored at
`./your_container.sif`.

```{r apptainer, eval=FALSE}
# To pull the docker image 
$ apptainer pull your_container.sif docker://rocker/geospatial:latest
# once the sif is downloaded, get into an interactive shell 
$ apptainer shell your_container.sif
# Inside the interactive shell, one can install INLA on a personal library path
$ R --verbose 
# Now, in a R environment
> options(repos = c(
    INLA = 'https://inla.r-inla-download.org/R/testing',
    CRAN = 'https://cloud.r-project.org'))
> install.packages("INLA") 
# One may be asked to create a personal library path.
> install.packages("inlabru") 
# quit R
> q()
```

3. Put the following in your job script to execute an R file non-interactively.

```{r exec, eval=FALSE}
# To execute an R file with apptainer
$ apptainer exec ./your_container.sif Rscript --no-restore --no-save --verbose file_to_run.R
```

Note: Use `inla.setOption(num.threads = ncpu)` to limit the number of
threads (ncpu) since INLA does not set it automatically. The `ncpu`
also directly links to RAM requirements.
