---
title: "Conditional sampling from a fitted model"
author: "Haavard Rue `(hrue@r-inla.org)`"
date: "Aug 19th 2017"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Conditional sampling from a fitted model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
---

```{r setup, include=FALSE}
set.seed(123)
library(INLA)
if (file.exists("myinit.R")) source("myinit.R")
inla.setOption(num.threads="1:1")
inla.setOption(smtp="taucs")
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
knitr::opts_chunk$set(fig.path="figures/conditional-sampling/")
```
This short note describe how to do conditional sampling from a fitted
model, where we want to condition on new events. An easy example which
motivated this note, is the following.
```{r}
n = 50
grp.len = 5
x = rnorm(n)
g = rnorm(grp.len)
grp = rep(1:grp.len, each = n %/% grp.len)
y = 1 + x + g[grp] + rnorm(n, sd = 0.01)
```
and the task is to provide the posterior a new observation, condition
on that the random effect 'g' is zero. Although it is possible to get
this directly (an exercise for the reader), in general, we will show that
it is easier to compute this using Monte Carlo sampling and the function
\verb|inla.posterior.sample()|. With a little hack. 

To fit the model, we do
```{r}
r = inla(y ~ x + f(grp, model = "iid"),
         data = data.frame(y, x, grp), 
         control.compute = list(config=TRUE)) 
```
adding \verb|config=TRUE| to prepare for the use of \verb|inla.posterior.sample()|.

The trick, is to add the conditioning as constraints, like
$$
A x = e
$$
where $x$ is the latent field. In our case, it is simply `t grp.len`
constraints, $\{g_i=0\}$ for
$i$ from 1 to `r grp.len`.
To do this, we need to create the matrix $A$, and vector $e$, and add
it to the list of contraints (if any). To do this, we need to know the
index of the \verb|g|-term, which we find here
```{r}
r$misc$configs$contents
```
so we have both the start index and the length of this vector. So we
create a constraint $Ax=0$ to represent 
constraints, $\{g_i=0\}$ for
$i$ from 1 to `r grp.len`.
```{r}
m = sum(r$misc$configs$contents$length)
grp.idx = which(r$misc$configs$contents$tag == "grp")
grp.len = r$misc$configs$contents$length[grp.idx]
A = matrix(0, grp.len, m)
e = matrix(0, grp.len, 1)
for(i in 1:grp.len) {
    A[i, r$misc$configs$contents$start[grp.idx] + i - 1] = 1
    e[i] = 0
}
```
Then we need to append it to the existing one if any, or to add a new one.
```{r}
constr = r$misc$configs$constr
if (is.null(constr)) {
    ## nothing there from before
    r$misc$configs$constr = list(
        nc = dim(A)[2],
        A = A,
        e = e)
} else {
    ## create a new one
    r$misc$configs$constr = list(
        nc = constr$nc + dim(A)[2],
        A = rbind(constr$A, A),
        e = rbind(constr$e, e))
}
```
When we now do
```{r}
xx = inla.posterior.sample(1000, r)
```
we additional condition on $Ax=e$, but \verb|g| would still appear in
the list of samples as zero or something very close to it.  The rest
should be straight forward.

PS: Note that you can add any linear combination of $x$ as a
constraint, you just do not want to many, as the cost is quadratic the
number of constraints. If the constraints simply set variables to
zero, you can also add a large number to $Q_{i,i}$, which you will
find in \verb|r$misc$configs|, one for each configuration.
