---
title: "Approximating joint marginals in R-INLA"
author: "Cristian Chiuchiolo and Haavard Rue"
date: "KAUST, Aug 2020"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Joint marginals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8} 
bibliography: jmarginal.bib
---

```{r setup, include=FALSE}
set.seed(1234)
library(INLA)
if (file.exists("myinit.R")) source("myinit.R")
inla.setOption(num.threads="1:1")
inla.setOption(smtp="taucs")
library(ggplot2)
library(sn)
library(knitr)
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(size='small'
               , cache=FALSE
               , cache.path='cache/'
               , comment=NA
               , warning=FALSE
               , message=FALSE
               , fig.align='center'
               , fig.path='figures/jmarginal/'
               , fig.pos='H'
               , background='#ffffff'
               , results='hold'
               , eval=TRUE)
```

## Introduction

The primary use of `R-INLA` is to approximate univariate marginals of
the latent field, so we can compute their marginal summaries and
densities. In applications, we sometimes need more than this, as we
are interested also in statistics which involve more several of the
components of the latent field, and/or, a joint posterior
approximation to a subset of the latent field.

The way around this issue, have earlier resolved to stochastic
simulation, using the function `inla.posterior.sample`. This function
samples from joint approximation to the full posterior which `INLA`
construct, but do so for the whole latent field. Using these samples,
we can compute the density of the relevant statistics and/or use
standard methods to represent a joint marginal.

This vignette introduces a new tool which computes a deterministic
approximation to the joint marginal for a subset of the latent field
using `R-INLA`. This approximation is explicitly available, and
constructed using skew-normal marginals and a Gaussian copula, hence
restricted to a joint approximation of a modest dimension.

The key specification is using an argument `selection` in the
`inla()`-call, which defines the subset, and then the joint marginal
approximation is made available in `result$selection`.

## Theory reference

For any \emph{Latent Gaussian Model} with data $\boldsymbol{y}$ and
set of parameters $(\boldsymbol{x}, \boldsymbol{\theta})$ with
$\boldsymbol{x}$ being the latent field and $\boldsymbol{\theta}$ the
hyperparameters, the resulting joint posterior approximation is stated
as

\begin{equation}
\tilde{\pi}(\boldsymbol{x},\boldsymbol{\theta}|\boldsymbol{y}) \propto \sum_k \tilde{\pi}_G (\boldsymbol{x}|\boldsymbol{\theta},\boldsymbol{y}) \tilde{\pi}(\boldsymbol{\theta}_k|\boldsymbol{y})  \Delta_k
\label{approx_joint}
\end{equation}

where $\tilde{\pi}_G
(\boldsymbol{x}|\boldsymbol{\theta},\boldsymbol{y})$ is the Gaussian
approximation. This expression recalls a Gaussian mixture distribution
with weights $\tilde{\pi}(\boldsymbol{\theta}_k|\boldsymbol{y})
\Delta_k$ obtained in the grid exploration of the hyperparameter
posterior marginals. For more insights, we suggest checking sources
like @inla, @thiago, @marta, or the recent review by @sara2019. The
Gaussian approximation used in \eqref{approx_joint} is both mean and
skewness corrected since it exploits Skew-Normal marginals of the
latent field into a Gaussian copula structure (see @egil for details).
These corrections are available in `inla.posterior.sample` as

\begin{verbatim}
inla.posterior.sample(..., skew.corr = TRUE)
\end{verbatim}

## First example

We will illustrate this new feature using a simple example.
```{r}
n = 100
x = rnorm(n, mean = 1, sd = 0.3)
xx = rnorm(n, mean = -2, sd = 0.3)
y = rpois(n, lambda = exp(x + xx))
r = inla(y ~ 1 + x + xx,
         data = data.frame(y, x, xx), 
	     family = "poisson")
```
Let us compute the joint marginal for the effect of `x`, `xx` and the
intercept. This is specified using the argument `selection`, which is
a named list of indices to select. Names are those given the formula,
plus standard names like `(Intercept)`, `Predictor` and `APredictor`.
So that
```{r}
selection = list(Predictor = 3:4, x = 1, xx = 1)
```
say that we want the joint marginal for the $3^{rd}$ and $4^{th}$ element of
`Predictor` and the first element of `x` and `xx`. (Well, `x` and `xx`
only have one element, so then there is not else we can do in this
case.)

If we pass `selection`, then we have to rerun `inla()` as
```{r}
rs = inla(y ~ 1 + x + xx,
          data = data.frame(y, x, xx), 
	      family = "poisson",
          control.compute = list(return.marginals.predictor = TRUE), 
	      control.predictor = list(compute = TRUE),
		  selection = selection)
```		 
and we obtain
```{r}
#summary(rs$selection)
print(rs$selection)
```
The Gaussian copula is given by the `mean` and the `cov.matrix`
objects, while the Skew-Normal marginals are given implicitly using
the marginal mean and variance in the Gaussian copula and the listed
skewness. Moreover the respective Skew-Normal mapping parameters for
the marginals $(\xi, \omega, \alpha)$ are provided in the object
'marginal.sn.par'. The `names` are given as a separate entry instead
of naming each individual result, to save some storage.

There are utility functions to sample and evaluate samples from this
joint marginal, similar to `inla.posterior.sample` and
`inla.posterior.sample.eval`. 
```{r}
ns = 10000
xs = inla.rjmarginal(ns, rs) ## or rs$selection
str(xs)
```
whose output is a matrix where each row contains the samples for the
variable in each column
```{r}
pairs(t(xs$samples[, 1:3000]), cex = 0.1)
```
We can compare the approximation of `Predictor:3` to the one computed
by the `inla()` call, 
```{r}
hist(xs$samples["Predictor:3",], n = 300, prob = TRUE, 
     main = 'Histogram of Predictor:3', xlab = 'Samples 
     from the joint and linear predictor marginal (black straight line)')
lines(inla.smarginal(rs$marginals.linear.predictor[[3]]), lwd = 3)
```
These marginals are not *exactly* the same (as they are different
approximations), but should be very similar. 

## Deterministic Joint approximation

As a conclusion to this vignette we show an additional joint posterior
related tool. The following INLA function computes a deterministic
approximation for the joint posterior sampler and must be considered
experimental. The function is strictly related to the selection type
INLA setting. Deterministic posterior marginals for the previous
example can be obtained as follows

```{r}
dxs = inla.1djmarginal(jmarginal = rs$selection)
str(dxs)
```

The output is a list with all computed marginals and a matrix summary
output in INLA style. Marginal can be accessed and plotted by using
the respective names in the selection

```{r}
ggplot(data = data.frame(y = xs$samples["Predictor:3",]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs$`Predictor:3`), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for Predictor:3"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = xs$samples["Predictor:4",]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs$`Predictor:4`), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for Predictor:4"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = xs$samples["x:1",]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs$`x:1`), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for x:1"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = xs$samples["xx:1",]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs$`xx:1`), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for xx:1"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red"))  
```

Here we compare the deterministic marginals with their sampling
version. They are quite accurate and can provide more informations.
Indeed, a complete summary based on these deterministic results is
achievable with a personalized `summary` function

```{r}
summary(rs$selection)
```

where posterior estimates and quantiles are computed for all the
selected marginals. Along the same line, we can easily compute
multiple deterministic linear combinations through the function
`inla.tjmarginal`and a matrix object A with the respective indexes

```{r}
A = matrix(c(1,1,0,0,0,0,1,1), nrow = 2, ncol = 4, byrow = T)
rownames(A) <- c("Test1", "Test2")
A
```

We define two linear combinations: `Predictor:3+Predictor:4` and
`x:1+xx:1` respectively. Then we can use the cited function which has
the same class of `selection` object

```{r}
m = inla.tjmarginal(jmarginal = rs, A = A)
m
class(m)

```

```{r}
dxs.lin = inla.1djmarginal(jmarginal = m)
str(dxs.lin)

fun1 <- function(...) {Predictor[1]+Predictor[2]}
fun2 <- function(...) {x+xx}

xs.lin1 = inla.rjmarginal.eval(fun1, xs)
xs.lin2 = inla.rjmarginal.eval(fun2, xs)

ggplot(data = data.frame(y = xs.lin1[1, ]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs.lin$Test1), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for Lin1"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = xs.lin2[1, ]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(dxs.lin$Test2), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for Lin2"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 
```

and accomplish the job with summaries

```{r}
summary(m)
```

Transformations of the marginal terms or linear combinations are
possible as well. We just need to use `inla.tmarginal` as follows

```{r}
fun.exp <- function(x) exp(x)

fun5 <- function(...) {exp(x)}
fun6 <- function(...) {exp(xx)}
fun7 <- function(...) {exp(x+xx)}

tdx <- inla.tmarginal(fun = fun.exp, marginal = dxs$`x:1`)
tdxx <- inla.tmarginal(fun = fun.exp, marginal = dxs$`xx:1`)
tdx.lin <- inla.tmarginal(fun = fun.exp, marginal = dxs.lin$Test2)

tx = inla.rjmarginal.eval(fun5, xs)
txx = inla.rjmarginal.eval(fun6, xs)
tx.lin = inla.rjmarginal.eval(fun7, xs)

ggplot(data = data.frame(y = tx[1, ]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(tdx), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for exp(x:1)"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = txx[1, ]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(tdxx), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for exp(xx:1)"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 

ggplot(data = data.frame(y = tx.lin[1, ]), aes(y, after_stat(density), colour = "True")) +
  stat_density(alpha = .1) +
  geom_line(data = as.data.frame(tdx.lin), aes(x = x, y = y, colour = "Deterministic"))+
  labs(title= '"Marginal Results for exp(x:1+xx:1)"', x='x', y='Density') +
  scale_colour_manual("", 
                      breaks = c("True", "Deterministic"),
                      values = c("black", "red")) 
```

Summaries for all marginal transformations can be obtained through
`inla.zmarginal`

```{r}
expx = inla.zmarginal(marginal = tdx, silent = TRUE)
expxx = inla.zmarginal(marginal = tdxx, silent = TRUE)
expx.lin = inla.zmarginal(marginal = tdx.lin, silent = TRUE)

exp.summaries = rbind(expx, expxx, expx.lin)
exp.summaries
```

## References
