---
title: "SPDE how to"
author: "Elias T. Krainski"
date: "created: March 31, 2016, last update `r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  \VignetteIndexEntry{SPDEhowto}
  \VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# A short introduction on how to fit a **SPDE** model using **INLA**

## Introduction 

This document ilustrates how to do a 
geostatistical fully Bayesian analysis
through the **S**tochastic **P**artial 
**D**iferential **E**quation approach 
<http://onlinelibrary.wiley.com/doi/10.1111/j.1467-9868.2011.00777.x/full>
using the **I**ntegrated **N**ested 
**L**aplace **A**proximation, 
<http://onlinelibrary.wiley.com/doi/10.1111/j.1467-9868.2008.00700.x/full>
implementation in the package available at 
<http://www.r-inla.org>.


```{r, include=FALSE}
library(knitr);  library(INLA);  library(fields) 
set.seed(1) 
```

## Simulating some data


Define some random **Locations** 
and the Random Field (RF) 
**covariance** matrix, considering
exponential correlation function: 

```{r, tidy=FALSE}
n = 200  ## number of location points
coo = matrix(runif(2*n), n) ## location points
k <- 10;   s2s <- 0.7 ## RF parameters
R <- s2s*exp(-k*as.matrix(dist(coo))) ## covariance matrix
```

Draw a **RF** sample: a multivariate Normal realization 
  
```{r, tidy=FALSE}
s <- drop(rnorm(n)%*%chol(R)) ## one RF realization
```

Adding a **covariate** effect and a noise

```{r,tidy=FALSE}
x <- runif(n)  ## covariate
beta <- 1:2  ## regression coefficients
lin.pred <- beta[1] + beta[2]*x + s ## linear predictor
s2e <- 0.3 ## error variance (nugget)
y <- lin.pred + rnorm(n, 0, sqrt(s2e))  ## the outcome
```

## Model fitting: steps

*  __Mesh__:
a triangulation to discretize the 
random field (RF) at 'm' nodes. 

```{r, tidy=FALSE, fig.width=5, fig.height=5.1, out.width='0.57\\textwidth', fig.align='center'}
mesh <- inla.mesh.2d( ## 2D mesh creator
  loc=coo, ## provided locations 
  ## works if only domain or boundary are provided
  max.edge=c(1/k, 3/k), ## maximum edge length (inner, outer): mandatory 
  offset=c(1/k, 5/k), ## outer extension
  cutoff=0.1/k) ## good to have >0
### visualize it
par(mar=c(0,0,1,0))
plot(mesh, asp=1) ## plot the mesh
points(coo, col='red') ## add the points 
```

A little *warning* about the mesh. 
The additional triangles outer domain 
is to avoid boundary effects. 
Is good to have aproximately 
isosceles triangles. 
And, to avoid tiny triangles. 
We need to have edges lengths 
of the inner mesh triangles 
less than the range of the process. 
Of course, if it is too small, there 
might not be any spatial effect. 

* __Define the__ _n x m_ __projector matrix__ to project 
 the process at the mesh nodes to locations 

```{r, projector, tidy=FALSE}
image(A <- inla.spde.make.A( ## projector creator
    mesh=mesh, ## provide the mesh
    loc=coo) ### locations where to project the field
    ) ## an 'n' by 'm' projector matrix
```

* __Build the SPDE model__ on the mesh.
Set $\alpha=3/2$ to build the precision structure for 
an Exponential correlation function

```{r,tidy=FALSE}
spde <- inla.spde2.matern( ## precision components creator
    mesh=mesh, ## mesh supplied
    alpha=1.5) ## smoothness parameter
```

* __Create a data stack__ to organize the data. 
This is a way to allow models with complex 
linear predictors. In our case, we have a SPDE 
model defined on _m_ nodes. It must be combined 
with the covariate (and the intercept) 
effect at _n_ locations. 
We do it using different projector matrices. 

```{r, tidy=FALSE}
stk.e <- inla.stack( ## stack creator
  data=list(y=y),  ## response
  effects=list(## two elements:
    data.frame(b0=1, x=x), ## regressor part
    s=1:spde$n.spde),  ## RF index
  A=list(## projector list of each effect
    1, ## for the covariates
    A), ## for the RF
  tag='est') ## tag
``` 

* __Fit__ the posterior marginal distributions
  for all model parameters 

```{r, tidy=FALSE, message=FALSE, warning=FALSE}
formula <- y ~ 0 + b0 + x + ## fixed part
  f(s, model=spde) ## RF term
res <- inla( ## main function in INLA package
  formula, ## model formula
  data=inla.stack.data(stk.e), ## dataset
  control.predictor=list( ## inform projector needed in SPDE models
    A = inla.stack.A(stk.e))) ## projector from the stack data
```


## Posterior marginal distributions - PMDs

Summary of the regression coefficients PMDs

```{r}
round(res$summary.fixed, 4) 
```
We have to transform the precision PMD to have 
the variance PMD. It can be done and visialized by

```{r, tidy=FALSE, fig.width=3.5, fig.height=3.5, out.width='0.35\\textwidth', fig.align='center', fig.pos='h'}
m.prec <- res$marginals.hyperpar$'Precision for the Gaussian observations' ## the marginal
post.s2e <- inla.tmarginal(## function to compute a tranformation 
  function(x) 1/x, ## consider the inverse transformation
  m.prec) ## of this marginal
### visualize it
par(mar=c(2.5,2.5,0.1,0.1), mgp=c(1.7,0.5,0))
plot(post.s2e, type='l', ylab='Density', 
     xlab=expression(sigma[e]^2))
abline(v=s2e, col=2) ## add value used to generate the data
```

The SPDE approach uses a local variance, $\tau^2$, 
such that $\sigma_{s}^2=1/(2\pi\kappa^2\tau^2)$. 
On \textbf{\textsf{INLA}} we work log($\tau^2$)
and log($\kappa$). So, especially for 
$\sigma_{s}^2$, we have to do 
an additional computation. 
The PMDs for all RF parameters on 
user scale are computed by

```{r} 
rf <- inla.spde.result( ## function to compute the 'interpretable' parameters
    inla=res, ## the inla() output
    name='s', ## name of RF index set
    spde=spde, ## SPDE model object
    do.transf=TRUE) ## to user scale
``` 

It can be visualized by 

```{r, fig.width=7.5, fig.height=2.5, out.width='0.99\\textwidth', out.height='0.33\\textwidth', fig.align='center', fig.pos='h', fig.keep='last'}
par(mfrow=c(1,3), mar=c(3,3,0.3,0.3), mgp=c(2,0.5,0))
plot(rf$marginals.var[[1]], ty='l', 
     xlab=expression(sigma[s]^2), yla='Density')
abline(v=s2s, col=2) ## add the true value
plot(rf$marginals.kap[[1]], type='l',
     xlab=expression(kappa), ylab='Density')
abline(v=k, col=2) ## add the true value
plot(rf$marginals.range[[1]], type='l', 
     xlab='range nominal', ylab='Density')
abline(v=sqrt(0.5 * 8)/k, col=2) ## add the 'true' value
```

## Projection on a grid / visualization

An interesting result is the map of the RF on a grid. 
The simplest way to have it is by projection. 
We just have to define the projector matrix 
and project, for example, the posterior 
mean and posterior standard deviation on the grid. 

```{r}
gproj <- inla.mesh.projector( ## projector builder
  mesh, ## mesh used to define the model
  xlim=0:1, ylim=0:1, ## limits where to create the grid
  dims=c(300,300)) ## grid dimension
## project the mean and the SD
g.mean <- inla.mesh.project(gproj, res$summary.random$s$mean)
g.sd <- inla.mesh.project(gproj, res$summary.random$s$sd)
```

We can visualize it by 
```{r, tidy=FALSE, fig.width=9, fig.height=5, out.width='0.97\\textwidth', fig.align='center', fig.pos='h'}
par(mfrow=c(1,2), mar=c(0,0,1,0))
require(fields)
image.plot(g.mean, asp=1, main='RF posterior mean', axes=FALSE, horizontal=TRUE)
image.plot(g.sd, asp=1, main='RF posterior SD', axes=FALSE, horizontal=TRUE)
```

## Prediction 

Define the set of target locations, 
the corresponding projector 
matrix and covariate values at target locations 

```{r}
tcoo <- rbind(c(0.3,0.3), c(0.5,0.5), c(0.7,0.7))
dim(Ap <- inla.spde.make.A(mesh=mesh, loc=tcoo)) 
x0 <- c(0.5, 0.5, 0.5)
```

To do a fully Bayesian analysis, 
we include the target locations on the 
estimation process by assigning NA 
for the response at these locations. 
Defining the prediction stack 

```{r}
stk.pred <- inla.stack(
  tag='pred', ## will be used to collect the posterior marginals
  data=list(y=NA), ## response set as NA
  effects=list(
    data.frame(x=x0, b0=1), ## covariate scenario
    s=1:spde$n.spde), ## same as before
  A=list(1, Ap)) ## covariate and target locations field projectors
``` 

Fit the model again with the full stack 

```{r}
stk.full <- inla.stack(stk.e, stk.pred) ## join both data
p.res <- inla(
  formula, data=inla.stack.data(stk.full), ## using the full data 
  control.predictor=list(compute=TRUE, ## compute the predictor
                         A=inla.stack.A(stk.full)), ## from full data
  control.mode=list(theta=res$mode$theta)) ## use the mode previously found
``` 

Get the prediction data index and collect the 
linear predictor PMDs  to work with 

```{r}
pred.ind <- inla.stack.index( ## stack index extractor function
  stk.full, ## the data stack to be considered
  tag='pred' ## which part of the data to look at
  )$data ## which elements to collect
ypost <- p.res$marginals.fitted.values[pred.ind]
```

Visualize with commands bellow

```{r, tidy=FALSE, fig.width=8, fig.height=3, out.width='0.99\\textwidth', out.height='0.35\\textwidth', fig.pos='h', fig.keep='last'}
xyl <- apply(Reduce('rbind', ypost), 2, range)
par(mfrow=c(1,3), mar=c(3,3,2,1), mgp=c(2,1,0))
for (j in 1:3) 
  plot(ypost[[j]], type='l', xlim=xyl[,1], ylim=xyl[,2], 
       xlab='y', ylab='density', main=paste0('y', j))
``` 

We have already used the *inla.tmarginal()* function. 
There are some other functions to work with 
marginal distributions which may be usefull as well: 

```{r}
apropos('marginal')
```

Playing with the posterior marginal for the first target location
```{r}
inla.qmarginal(c(0.15, 0.7), ypost[[1]]) ## quantiles
inla.emarginal(function(x) x^2, ypost[[1]]) - ## E(y^2) -
  inla.emarginal(function(x) x, ypost[[1]])^2 ## E(y)^2 = variance
inla.pmarginal(inla.qmarginal(0.3, ypost[[1]]), ypost[[1]]) 
inla.zmarginal(ypost[[1]]) ## posterior summary
```
