---
title: "SPDE example with multiple kind of outcomes:"
subtitle: "Joint modeling with `mdata`, `inla.surv` and counts with exposure."
author: "Elias T Krainski"
date: "2023, March"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{SPDE example with multiple kind of outcomes.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: spdeshort.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  fig.width = 10,
  fig.height = 7, 
  out.width = "69%",
  fig.align = "center"
)
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
```

# Abstract 

The **INLA** package supports different kinds of data likelihoods,
including the `inla.surv`, which is a generalization of the
`Surv` from the **survival** package, and `mdata`.
It also support joint modeling different kind of outcomes.
In this vignette we illustrate the case when we have
three outcomes, one of each type, and model them jointly.

# Introduction 

Usually correlations between outcomes are modeled by
sharing common model terms in the linear predictor for each one.
These terms can be covariates, random effects or sums of these.
It is possible to have cases when each likelihood for an
observation unit depend on more than one quantity, 
for example, one likelihood is to model the survival time, 
which has the additional information about censoring, 
and other likelihood is written as a function of several quantities, 
as we will see in the example.

In this vignette we consider a joint model for three outcomes,
simulate data from this model and illustrate how to fit the 
joint model accounting for shared terms.
The linear predictor for each outcome includes covariates and 
a spatial random effect modeled using the SPDE approach.

We consider a scenario where the first outcome is known at an aggregated level,
in a setting were we model it considering a kind of likelihood designed in
**INLA** to account for all the information available with respect to the aggregation.
The second outcome is considered to be observed censored,
so that we have to consider the observed value and the censoring information.
The third outcome is just the usual kind of data: one scalar per data unit.

## The spatial domain 

The SPDE models available in **INLA**
could be applied for processes in $\mathbb{R}^d$
or $\mathbb{S}^d$, for $d=1,2$, and the theory in @lindgrenRL2011 
include higher dimentional domains.
We consider a spatial domain, $D \in \mathbb{R}^2$
to keep the example simple.

Without loss off generality we consider $D$ as a rectangle defined as
```{r domain}
bb <- rbind(c(0, 20), c(0, 12))
```

Next, we store the length of each side and its average for later use
```{r rxy}
rxy <- apply(bb, 1, diff)
b <- mean(rxy)
```

We sample a set of $n$ locations,
$l_1, ..., l_n$, in this domain, $l_i \in D$, 
completely at random as follows:
```{r loc}
n <- 500
loc <- cbind(
    runif(n, bb[1, 1], bb[1, 2]),
    runif(n, bb[2, 1], bb[2, 2]))
```

## A spatial field

Instead of simulating from a model to play as a spatial smooth term, 
we define a smooth function on this domain as follows
```{r sfn}
sfn <- function(a, b)
    sin(a + b) + cos(a - b)
```

This function is evaluated at the $n$ locations and visualized with
```{r sloc}
u <- sfn(2 * pi * loc[, 1] / rxy[1],
         2 * pi * loc[, 2] / rxy[2])
summary(u)
par(mar = c(2, 2, 0, 0), mgp = c(1.5, 0.5, 0), las = 1)
plot(loc, pch = 19, cex = 0.5 + (2 + u)/4, xlab = "", ylab = "")
```

## Spatial sub-domains 

In order to define the first outcome,
let us consider that the domain is being divided into a set of sub-regions, 
$r_1, r_2, ..., r_g$, 
so that $r_i \cap r_i = \oslash$ and $\cup_{i=1}^g r_j = D$.

For simplicity of coding, but without loss of generality,
we assume that these small sub-regions to be pixels based in
the following set of centers:
```{r centers}
h <- 2 ## size of the side of each sub-region
group.ce <- as.matrix(expand.grid(
    seq(bb[1, 1] + h/2, bb[1, 2], h),
    seq(bb[2, 1] + h/2, bb[2, 2], h)))
(ng <- nrow(group.ce))
```

We can avoid the need of dealing with the general
spatial objects and operations available in **R**
and identify each location $l_i$ to each sub-region $r_j$
with an integer vector as
```{r groupid}
group.id <- ceiling(loc[,1]/h)+ceiling(rxy[1]/h) *
  (ceiling(loc[,2]/h)-1)
```

This will be used as a grouping index, so that each group is made of the
observations falling in each of the pixels.
We will visualize the locations with color depending on this grouping later.

## Covariates

Let us consider a set of three covariates for each location as 
```{r xxx}
xxx <- cbind(x1 = runif(n), x2 = runif(n), x3 = runif(n))
```

We average the covariates by group as we will consider that
we have data at the group level to model the first outcome.
```{r x.g}
xxx.g <- aggregate(xxx, list(g = group.id), mean)
str(xxx.g)
```


# Outcomes

We will consider three outcomes.

### First outcome: aggregated Gaussian

We assume that one of the outcomes is a continuous random variable,
to be considered as Gaussian, with unknown mean and variance,
and a known scaling factor, $s$, such that 
\[ 
y_i \sim \textrm{N}(\mu_i, s_i\sigma^2).
\]

The first outcome being simulated considering that the
expected value as $\eta_1 = \{\mu_1, ..., \mu_n\}$ as
\[ 
\eta_1 = \mathbf{X}\beta_1 + \mathbf{u}.
\]
This is being evaluated considering the covariate and random effect averaged
per group:
```{r y}
betas1 <- c(5, 0, -3, 3) 
u.g <- tapply(u, group.id, mean)
eta.y.g <- drop(cbind(1, as.matrix(xxx.g[, 2:ncol(xxx.g)])) %*% betas1) + u.g
```

We will simulate the scaling factor for each observation with
```{r s}
s <- rgamma(n, 7, 3)
summary(s)
```

```{r ysim}
sigma.y <- 1
y <- rnorm(n, eta.y.g[group.id], sqrt(sigma.y/s))
summary(y)
```

We simulated $y_i$ for each location $l_i$.
However, we consider that observations of this variable are
available only at the aggregated level, at each of the sub-regions $r_j$.
Furthermore, for each sub-region $r_j$, we know 

 - how many observations were taken in each sub-region, $n_j = \sum_{i=1}^n I(l_i \in r_j)$
 - the weighted average $\overline{y}_j$ = $(n_j)^{-1}\sum_{i,l_i\in r_j} s_iy_i$,
 - the sum o the scaling factors, $m_j$ = $\sum_{i,l_i\in r_j} s_i$, 
 - the half of the sum of the log of the scaling factors, $\frac{1}{2}\sum_{i,l_i\in r_j} \log(s_i)$,
 - and the sample variance within each unit, defined as
 $v_j$ = $\frac{1}{n_j}\sum_{i,l_i\in r_j}^n(s_i y_i^2-\overline{y}_j^2)$

The function `inla.agaussian` computes these five statistics, 
given $y_i$ and $s_i$ for each group.
We use this function to compute these statistics, as follows 
```{r ag, warning}
library(INLA)
agg <- lapply(1:ng, function(g) {
    ig <- which(group.id==g)
    if(length(ig)>0) 
        return(inla.agaussian(y[ig], s[ig]))
    return(inla.mdata(list(NA, 0, 1, 1, NA))) ### a deal with missing
})
str(YY <- Reduce('rbind', lapply(agg, unlist))) ## five columns matrix
```

## Second outcome: Survival

One can fit survival modes using **INLA** considering several available 
likelihoods. One is the `weibullsurv` and we will now consider
that we have an outcome that follows a Weibull distribution, 
considering the following parametrization
\[
f_W(w) = \alpha w^{\alpha - 1} \lambda^\alpha \textrm{e}^{-(\lambda w)^\alpha}.
\]
We assume that $\lambda = \exp(\eta_2)$, 
where $\eta_2$ is the linear predictor and is defined as
\[ 
\eta_2 = \mathbf{X}\beta + \gamma_1\mathbf{u}
\]
and we consider the following code to sample from a Weibull distribution
```{r w0}
betas2 <- c(1, -1, 0, 1)
alpha <- 2
gamma1 <- 0.5
lambdaw <- exp(cbind(1, xxx) %*% betas2 + gamma1 * u)
we0 <- rweibull(n, shape = alpha, scale = 1/lambdaw)
summary(we0)
```

However, suppose we do not observe the survival times for some of the observations.
```{r event}
summary(u.ev <- runif(n, 0.3, 1)) ## censoring factor
table(event <- rbinom(n, 1, 0.5)) ## censored (=0) or event (=1)
summary(we <- ifelse(event == 1, we0, we0 * u.ev)) ## censored outcome
```

## Third outcome: Count under exposure

We consider that we have different exposure at each location
```{r ee}
summary(ee <- rgamma(n, 10, 2))
``` 

We assume an outcome following a Poisson distribution with 
\[ 
o_i \sim \textrm{Poisson}(\lambda_i E_i)
\]
where $\lambda_i$ depends on the covariates and assuming that 
\[
\log{\lambda} = \eta_3 = \mathbf{X}\beta_e + \gamma_2\mathbf{u}
\]

This outcome is being simulated with
```{r po}
betas3 <- c(2, 1, -1, 0)
gamma2 <- -0.3
lambdap <- cbind(1, xxx) %*% betas3 + gamma2 * u
po <- rpois(n, exp(lambdap) * ee)
summary(po)
```

# Data model setup 

We now have to setup the objects in order to fit the model.

## SPDE model setup

We have to define a mesh over the domain. 
We do it considering the following code
```{r mesh}
mesh <- inla.mesh.2d(
    loc.domain = matrix(bb[c(1,3,3,1,1, 2,2,4,4,2)], ncol = 2), 
    max.edge = c(.05, .15) * b,
    offset = c(.1, .2) * b)
```

This mesh and the locations colored by the group defined for the first
outcome can be visualized with
```{r vmesh}
par(mar = c(0, 0, 1, 0), mgp = c(1.5, 0.5, 0))
plot(mesh, asp = 1, edge.color = gray(0.5))
points(loc, pch = 19, col = group.id, cex = 0.3 + (2 + u)/4)
```

The SPDE model is defined with
```{r spde}
spde <- inla.spde2.pcmatern(
    mesh,
    prior.range=c(1, 0.1),
    prior.sigma=c(1, 0.1))
```

## Linear predictor components

We will fit a joint model considering the linear predictor
with terms as how it was simulated.
The `inla.stack()` function in ment to me used
when working with SPDE models.
However, it also helps when considering multiple likelihood data
as detailed bellow.

One way to deal with multiple likelihoods in **INLA**
is defining `matrix` data for the outcome where each column
is used for each likelihood.
However, we do have more complex cases.
Therefore, **the way** to deal with more complex multiple likelihood data is to
use `list` in the left hand side of the `formula`.
This is the key point to implement this example.

The right hand side of the formula includes each term in the linear predictor.
When working with multiple likelihood, each term in each likelihood
should be named uniquely.
For example, if two likelihoods does includes the same covariate
and each one with a different coefficient,
each one has to have its own name.
Therefore, we will pass the covariate `x1` as `x1y`, 
for the first outcome, `x1w` for the second outcome, 
and `x1p` for the third outcome.
Similarly for the other ones, including the intercept.

In our example, the SPDE model is in the linear predictor for the first outcome,
a copy from this random effect in in the linear predictor for the second outcome,
and another copy of this term is in the linear predictor for the third outcome.

The model formula considering all of these details is defined as
```{r formula}
fff <- list(
  inla.mdata(Y),
  inla.surv(w, ev),
  o) ~ 0 +
  a1 + x1y + x2y + x3y + f(s, model = spde) +
  a2 + x1w + x2w + x3w + f(sc1, copy = "s", fixed = FALSE) +
  a3 + x1p + x2p + x3p + f(sc2, copy = "s", fixed = FALSE)
```

## Data stacks

The data has to be organized so it contains all the information needed,
that is all the outcomes, covariates and random effect indexes.
For the SPDE models it also has to include the projector matrices.

The main work of the `inla.stack()` function is to check and organize
outcomes, effects and projector matrices
in order to guarantee that the dimentions match.
Additionally, `inla.stack()` will deal with the problem of
arranging the outcomes to work properly with multiple
likelihoods, without us having to deal with the
completion of lines and columns with NA.
We just have to supply the data for each outcome and its linear predictor
in a separated `inla.stack()` call and then
it does the job properly when joining all the data stacks.

For the first outcome, we need to supply a five column matrix,
each column with each of the five statistics computed
for each of the `ng` groups, previously computed.
For the effects we define an intercept, 
each one of the covariates with names according to what the
`formula` specify and the index for the SPDE model.
The projector matrices is one for the fixed effect, which is just `1`,
and the one for the SPDE model.

The stack for the first outcome can be defined as
```{r stack1}
stackY <- inla.stack(
    tag = 'y',
    data = list(Y = YY),
    effects= list(
        data.frame(a1 = 1,
                   x1y = xxx.g$x1, 
                   x2y = xxx.g$x2, 
                   x3y = xxx.g$x3),
        s = 1:mesh$n),
    A=list(1,
           inla.spde.make.A(mesh, group.ce))
)
```

For the second outcome, we have to supply
two vectors to build the `inla.surv` data,
the covariates, with names according to the terms in the formula, 
and the projector matrices, the single `1` for the covariates and
the projector for the shared SPDE term.

The stack for the second outcome can be defined as
```{r stack2}
stackW <- inla.stack(
    tag = 'w', 
    data = list(w = we,
                ev = event), 
    effects = list(
        data.frame(a2 = 1,
                   x1w = xxx[, 1], 
                   x2w = xxx[, 2], 
                   x3w = xxx[, 3]),
        sc1 = 1:mesh$n),
    A = list(1,
             inla.spde.make.A(mesh, loc))
)
```

For the third outcome we have to supply the vector of counts
and the vector with the exposure. 
The effects and projector are similar with the previous others,
just that we have to account for its names.

The stack for the third outcome can be defined as
```{r stack3}
stackP <- inla.stack(
    tag = 'p', 
    data = list(o = po,
                E = ee),
    effects = list(
        data.frame(a3 = 1,
                   x1p = xxx[, 1], 
                   x2p = xxx[, 2], 
                   x3p = xxx[, 3]),
        sc2 = 1:mesh$n),
    A = list(1,
             inla.spde.make.A(mesh, loc))
)
```

Joining all the data into one data stack
```{r jstack}
stacked <- inla.stack(stackY, stackW, stackP)
```

# Fitting the model

We now supply this to the `inla()` function in order to fit the model
```{r fit}
result <- inla(
   formula = fff,
    family = c("agaussian", "weibullsurv", "poisson"),
    data = inla.stack.data(stacked),
    E = E, 
    control.predictor = list(
        A=inla.stack.A(stacked)),
    control.family = list(
        list(),
        list(variant = 1),
        list())
)
result$cpu
```

We now compare the true value for each 
fixed effect with its posterior summary.

```{r fixef}
round(cbind(
    true = c(betas1, betas2, betas3),
    result$summary.fix), 2)
```

For the hyper-parameters, we do have fitted one for the
`agaussian` likelihood, 
one for the `weibulsuv` and we have those for the 
SPDE model assumed for the spatial smooth effect 
(defined from the sine and cosine functions)
and the fitted parameters for this term shared 
from the first to the second and third outcomes.

```{r hpars}
round(cbind(true = c(1/sigma.y^2, alpha, NA, NA, gamma1, gamma2), 
            result$summary.hy), 2)

```

# References
