---
title: "SPDE one dimensional example"
author: "Elias T. Krainski"
date: 'created: April 2, 2016, last update `r Sys.Date()`'
output: pdf_document
vignette: |
  %\\VignetteIndexEntry{SPDE1d}   %\\VignetteEngine{knitr::knitr}   \usepackage[utf8]{inputenc}
---

# A **SPDE** model for one dimensional data 

## Introduction 

In this example we show row to analyse a time series 
of daily temperature using a one dimension SPDE model.
More details about it are on the paber at 
<https://www.jstatsoft.org/article/view/v063i19>

```{r settings, include=FALSE}
library(knitr);  library(INLA);  library(fields) 
opts_chunk$set(message=FALSE, warning=FALSE, tidy=FALSE,
fig.path='figures/SPDE1d')
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(0.1, 0.1, .1, .1))  # smaller margin on top and right
})
knit_hooks$set(mar3311 = function(before, options, envir) {
    if (before) par(mar = c(3, 3, 1, 1), mgp=c(2, 0.7, 0))  # smaller margin on top and right
})
```

## The data 

We consider the daily weather data available at 
<http://www.yr.no/>. 
We have the following set the URL for the daily data for 
Trondheim considerint in the last 13 months 
```{r url}
u0 <- paste0('http://www.yr.no/place/Norway/S%C3%B8r-Tr%C3%B8ndelag/',
             'Trondheim/Trondheim/detailed_statistics.html')
### browseURL(u0) ### to visit the web page 
```

One can read and extract the desired data table 
(the second one at the URL) using the **XML** package with
```{r xml, eval=FALSE}
require(XML)
d <- readHTMLTable(u0)[[2]]
```
However, it still need some pre-processing.

Without the **XML** package one can use 
```{r read}
d0 <- readLines(u0) ### read it as text
i <- grep('<tr>', d0) ### index for each table line
i <- i[i>grep('<tbody>', d0)[2]] ### select those for the second table
```

The desired data we would like to analyse is the 
minimum and maximum temperature. 
Commands to extract and pre-process these data
```{r data}
dates <- as.Date(d0[i+1], format='      <th>%b %d, %Y</th>')
tmed <- as.numeric(gsub('<td>', '', gsub('째</td>', '', d0[i+4])))
(n <- length(dates)) ### it is daily over last 13 months
```

Visualize it
```{r visualize, fig.width=10, fig.height=3}
pd <- pretty(c(dates, max(dates+30)), n=13)
par(mfrow=c(1,1), mar=c(3,3,0.5,2), mgp=c(2,.7,0), las=2, xaxs='i')
plot(dates, tmed, type='l', lwd=2,
     axes=FALSE, xlab='day', ylab='Temperature')
abline(h=0)
abline(h=3*(-8:9), v=pd, lty=3, col=gray(.5))
box()
axis(2, 3*(-8:9)); axis(4, 3*(-8:9))
axis(1, pd, months(pd, TRUE))
```

## Model fitting

*  __Mesh__:
in 1d it is a matter of chosing a set of knots, 
the order of the basis functions and the boundary. 
Choosing first order basis function and Neumann boundary.

```{r mesh}
coo <- as.numeric(dates-min(dates)) ## have numeric temporal coordinates
mesh <- inla.mesh.1d(loc=seq(min(coo), max(coo), by=7), ## knots (7 days)
                 boundary='neumann', degree=2)  ### boundary and basis function degree
```

* __Define the__ _n x m_ __projector matrix__ to project 
 the process at the mesh nodes to locations 

```{r, projector, fig.width=8, fig.height=2, small.mar=TRUE}
A <- inla.spde.make.A( ## projector creator
    mesh=mesh, ## provide the mesh
    loc=coo) ### locations where to project the field
dim(A) ## an 'n' by 'm' projector matrix
summary(rowSums(A)) ### each line sums up to one
summary(colSums(A)) ### 'how many' observations per knot
```

* __Build the SPDE model__ on the mesh.
Set $\alpha=2$ to build the precision 

```{r SPDE model}
spde <- inla.spde2.matern( ## precision components creator
    mesh=mesh, ## mesh supplied
    alpha=2) ## smoothness parameter
```

* __Create a data stack__ to organize the data. 
This is a way to allow models with complex 
linear predictors. 
In our case, we have a SPDE 
model defined on _m_ nodes. 
It must be combined 
with the covariate (and the intercept) 
effect at _n_ locations. 
We do it using different projector matrices. 

```{r data stack}
stk.e <- inla.stack( ## stack creator
  data=list(y=tmed),  ## response
  effects=list(## two elements:
    data.frame(b0=rep(1, n)), ## regressor part
    i=1:spde$n.spde),  ## RF index
  A=list(## projector list of each effect
    1, ## for the covariates
    A), ## for the RF
  tag='est') ## tag
``` 

* __Fit__ the posterior marginal distributions
  for all model parameters 

```{r fitting}
formula <- y ~ 0 + b0 + ## fixed part
  f(i, model=spde) ## RF term
res <- inla( ## main function in INLA package
  formula, ## model formula
  data=inla.stack.data(stk.e), ## dataset
  control.predictor=list( ## inform projector needed in SPDE models
    A = inla.stack.A(stk.e), compute=TRUE)) ## projector from the stack data
```

## Posterior marginal distributions - PMDs

Summary of the regression coefficients PMDs

```{r fixed}
round(res$summary.fixed, 4) 
```
We have to transform the precision PMD to have 
the variance PMD. It can be done and visialized by

```{r nugget, fig.width=4, fig.height=3, mar3311=TRUE}
m.prec <- res$marginals.hyperpar$'Precision for the Gaussian observations' ## the marginal
post.s2e <- inla.tmarginal(## function to compute a tranformation 
  function(x) 1/x, ## inverse transformation
  m.prec) ## marginal to be applied
plot(post.s2e, type='l', ylab='Density', 
     xlab=expression(sigma[e]^2))
```

The SPDE approach uses a local variance, $\tau^2$, 
such that $\sigma_{s}^2=1/(2\pi\kappa^2\tau^2)$. 
On \textbf{\textsf{INLA}} we work log($\tau^2$)
and log($\kappa$). So, especially for 
$\sigma_{s}^2$, we have to do 
an additional computation. 
The PMDs for all RF parameters on 
user scale are computed by

```{r rf} 
rf <- inla.spde.result( ## function to compute the 'interpretable' parameters
    inla=res, ## the inla() output
    name='i', ## name of RF index set
    spde=spde, ## SPDE model object
    do.transf=TRUE) ## to user scale
``` 

It can be visualized by 

```{r parameters, fig.width=10, fig.height=3}
par(mfrow=c(1,3), mar=c(3,3,0.3,0.3), mgp=c(2,0.5,0))
plot(rf$marginals.var[[1]], ty='l', 
     xlab=expression(sigma[s]^2), yla='Density')
plot(rf$marginals.kap[[1]], type='l',
     xlab=expression(kappa), ylab='Density')
plot(rf$marginals.range[[1]], type='l', 
     xlab='range nominal', ylab='Density')
```

## Predicted 

Visualize it with the commands bellow

```{r predicted, fig.width=10, fig.height=3}
par(mfrow=c(1,1), mar=c(3,3,0.3,2), mgp=c(2,0.5,0), las=2, xaxs='i')
id <- inla.stack.index(stk.e, tag='est')$data
plot(dates, tmed, type='l', axes=FALSE, ylab='Temperature')
for (j in 3:5)
  lines(dates, res$summary.fitted.values[id, j], lty=3)
box(); axis(2, 3*(-8:9)); axis(4, 3*(-8:9))
axis(1, pd, months(pd, T))
abline(h=0)
abline(h=3*(-8:9), v=pd, lty=3, col=gray(.5))
``` 

## Just a look to the rest of the data 

Pre-processing the maximum, minimum and normal temperature, 
the precipitation, and the average and maximum wind:
```{r other}
tmax <- as.numeric(gsub('<td>', '', gsub('째</td>', '', d0[i+2])))
tmin <- as.numeric(gsub('<td>', '', gsub('째</td>', '', d0[i+3])))
tnormal <- as.numeric(gsub('<td>', '', gsub('째</td>', '', d0[i+5])))
prec <- as.numeric(gsub('<td>', '', gsub('mm</td>', '', d0[i+6])))
wind <- as.numeric(gsub('<td>', '', gsub('m/s</td>', '', d0[i+10])))
wmax <- as.numeric(gsub('<td>', '', gsub('m/s</td>', '', d0[i+9])))
```

Visualize it
```{r visualize3, fig.width=10, fig.height=9}
par(mfrow=c(3,1), mar=c(0.1,3,0.1,2), mgp=c(2,.7,0), las=2, xaxs='i')
plot(dates, tmed, type='l', ylim=range(tmin, tmax, na.rm=TRUE), 
     axes=FALSE, xlab='', ylab='Temperature', col='green')
lines(dates, tmin, col='blue')
lines(dates, tmax, col='red')
lines(dates, tnormal)
legend(dates[which.min(tmin)], par()$usr[4], c('normal', 'max.', 'aver.', 'min.'), 
      col=1:4, lty=1, ncol=2, xjust=0.5, bty='n')
abline(h=5*(-5:6), v=pd, lty=3, col=gray(.5))
box(); axis(2, 5*(-5:6)); axis(4, 5*(-5:6))

plot(dates, prec, type='l', axes=FALSE, xlab='')
box(); axis(2); axis(4)
abline(v=pd, h=10*(1:4), lty=3, col=gray(0.5))

par(mar=c(3, 3, 0.1, 2), new=FALSE)
plot(dates, wind, type='l', axes=FALSE, xlab='',
     ylim=range(wind, wmax, na.rm=TRUE))
lines(dates, wmax, col=2)
box(); axis(2); axis(4)
abline(v=pd, h=5*(1:3), lty=3, col=gray(0.5))
axis(1, pd, months(pd, TRUE))
```

We can have a look at the difference between the daily mean temperature and 
the normal temperature. 

```{r anomalia, fig.width=10, fig.height=3.5}
par(mar=c(3, 3, 0.1, 2), mgp=c(2,0.7,0), las=2, xaxs='i')
plot(dates, tmed-tnormal, type='l', axes=FALSE, 
     xlab='', ylab='Deviation from the normal temperature')
box(); axis(2); axis(4)
abline(h=5*(-2:2), v=pd, lty=2, col=gray(0.5))
axis(1, pd, months(pd, TRUE))
```

