---
title: "inla.pc.vminf"
author: "Xiang Ye"
date: "2025-03-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(INLA)
source("inla.pc.vminf.R")
```


# inla.pc.dvminf

$0 \leq \kappa \leq 10$:

```{r}
k <- seq(0,10,0.01)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```


$0 \leq \kappa \leq 350$:

```{r}
k <- seq(0,350,0.01)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```


$350 < \kappa \leq 700$:

```{r}
k <- seq(351.01,700,0.01)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```


$\kappa > 700$:

```{r}
k <- seq(700.01,2000,0.01)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```




Around $\kappa=350$:

```{r}
k <- seq(300,400,0.1)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```


Around $\kappa=700$:

```{r}
k <- seq(600,800,0.1)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```


For $\kappa$ too large:

```{r}
k <- c(1e+308,2e+308,1e+309,1e+400)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
dens
dens.log
```


**Entire domain:**

```{r}
k <- seq(0,1000,0.01)
dens <- inla.pc.dvminf(k, lambda=1)
dens.log <- inla.pc.dvminf(k,lambda=1,log=TRUE)
par(mfrow=c(1,2))
plot(k,dens,type="l")
plot(k,dens.log,type="l")
```




# inla.pc.pvminf

$0 \leq \kappa \leq 700$:

```{r}
q.low <- seq(0,700,0.1)
q.high <- seq(700.1,1500,0.1)
prob.low <- inla.pc.pvminf(q.low, lambda=1)
prob.high <- inla.pc.pvminf(q.high, lambda=1)
prob.low.log <- inla.pc.pvminf(q.low, lambda=1,log=TRUE)
prob.high.log <- inla.pc.pvminf(q.high, lambda=1,log=TRUE)
par(mfrow=c(1,4))
plot(q.low,prob.low,type="l")
plot(q.high,prob.high,type="l")
plot(q.low,prob.low.log,type="l")
plot(q.high,prob.high.log,type="l")
```

For $\kappa$ too large:

```{r}
q <- c(1e+308,2e+308,1e+309,1e+400)
prob <- inla.pc.pvminf(q, lambda=1)
prob.log <- inla.pc.pvminf(q,lambda=1,log=TRUE)
prob
prob.log
prob==1
```


**Entire domain:**

```{r}
q.700 <- seq(600,800,0.01)
q <- seq(0,1000,0.1)
prob.700 <- inla.pc.pvminf(q.700, lambda=1)
prob <- inla.pc.pvminf(q, lambda=1)
par(mfrow=c(1,2))
plot(q.700,prob.700,type="l")
plot(q,prob,type="l")

```


# inla.pc.qvminf

**boundary values:**

```{r}
print(inla.pc.qvminf(0,lambda=1))
print(inla.pc.qvminf(1,lambda=1))
```
**small $\lambda$**:

```{r}
p <- seq(0,0.999,0.001)
quant.1 <- inla.pc.qvminf(p,lambda=1)
quant.10 <- inla.pc.qvminf(p,lambda=10)
quant.01 <- inla.pc.qvminf(p,lambda=0.1)
par(mfrow=c(1,3))
plot(p,quant.1,type="l")
plot(p,quant.10,type="l")
plot(p,quant.01,type="l")
```

```{r}
p <- seq(0.99,0.999,0.00001)
quant.1 <- inla.pc.qvminf(p,lambda=1)
quant.10 <- inla.pc.qvminf(p,lambda=10)
quant.01 <- inla.pc.qvminf(p,lambda=0.1)
par(mfrow=c(1,3))
plot(p,quant.1,type="l")
plot(p,quant.10,type="l")
plot(p,quant.01,type="l")
```


**large $\lambda$ with default len (2048L) and len=10000L**:

```{r}
p <- seq(0,0.999,0.001)
quant.2048L <- inla.pc.qvminf(p,lambda = 100)
quant.4096L <- inla.pc.qvminf(p,lambda = 100, len = 10000L)
par(mfrow=c(1,2))
plot(p,quant.2048L,type="l")
plot(p,quant.4096L,type="l")
# Both have warnings
```


# inla.pc.rvminf

```{r}
n <- 3000
```

```{r}
set.seed(1314)
ks.1 <- inla.pc.rvminf(n,lambda=1)
ks.10 <- inla.pc.rvminf(n,lambda=10)
ks.100 <- inla.pc.rvminf(n,lambda=100)
ks.1000 <- inla.pc.rvminf(n,lambda=1000)
par(mfrow=c(1,4))
hist(ks.1)
hist(ks.10)
hist(ks.100)
hist(ks.1000)
```


Log-scale:




```{r}
set.seed(520)
ks.10 <- inla.pc.rvminf(n,lambda=10)
log.k <- seq(0.01,25,0.01)
dens.10 <- inla.pc.dvminf(log.k,lambda=10)/(exp(log.k))
par(mfrow=c(1,2))
hist(log(ks.10))
plot(log.k, dens.10, type='l')
```

```{r}
set.seed(520)
ks.50 <- inla.pc.rvminf(n,lambda=50)
log.k <- seq(0.01,25,0.01)
dens.50 <- inla.pc.dvminf(log.k,lambda=50)/(exp(log.k))
par(mfrow=c(1,2))
hist(log(ks.50))
plot(log.k, dens.50, type='l')
```

```{r}
set.seed(520)
ks.100 <- inla.pc.rvminf(n,lambda=100)
log.k <- seq(0.01,25,0.01)
dens.100 <- inla.pc.dvminf(log.k,lambda=100)/(exp(log.k))
par(mfrow=c(1,2))
hist(log(ks.100))
plot(log.k, dens.100, type='l')
```

```{r}
set.seed(520)
ks.300 <- inla.pc.rvminf(n,lambda=300)
log.k <- seq(0.01,25,0.01)
dens.300 <- inla.pc.dvminf(log.k,lambda=300)/(exp(log.k))
par(mfrow=c(1,2))
hist(log(ks.300))
plot(log.k, dens.300, type='l')
```

```{r}
set.seed(520)
ks.1000 <- inla.pc.rvminf(n,lambda=1000)
log.k <- seq(0.01,60,0.01)
dens.1000 <- inla.pc.dvminf(log.k,lambda=1000)/(exp(log.k))
par(mfrow=c(1,2))
hist(log(ks.1000))
plot(log.k, dens.1000, type='l')
```








# additional:



```{r}
par(mfrow=c(1,2))
k <- seq(0,1000,0.1)
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=10),lambda=10),col="red")
abline(a=0,b=1)
k <- seq(600,800,0.1)
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=10),lambda=10),col="red")
abline(a=0,b=1)
```


$\lambda=1$:

```{r}
par(mfrow=c(1,2))
k <- seq(0,1e+25,1e+22)
res <- inla.pc.qvminf(inla.pc.pvminf(k,lambda=1),lambda=1)
plot(k,res,col="red")
abline(a=0,b=1)

k <- seq(0,1e+26,1e+23)
res <- inla.pc.qvminf(inla.pc.pvminf(k,lambda=1),lambda=1)
plot(k,res,col="red")
abline(a=0,b=1)
```

```{r}
par(mfrow=c(1,2))
k <- seq(0,1e+27,1e+24)
res <- inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=1,log=TRUE)),lambda=1)
plot(k,res,col="red")
abline(a=0,b=1)

k <- seq(0,1e+28,1e+25)
res <- inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=1,log=TRUE)),lambda=1)
plot(k,res,col="red")
abline(a=0,b=1)
```


$\lambda=0.1$:

```{r}
par(mfrow=c(1,3))
k <- seq(0,1e+24,1e+21)
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=0.1),lambda=0.1),col="red")
abline(a=0,b=1)

k <- seq(0,1e+25,1e+22)
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=0.1),lambda=0.1),col="red")
abline(a=0,b=1)

k <- seq(0,1e+26,1e+24)
plot(k,inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=0.1,log=TRUE)),lambda=0.1),col="red")
abline(a=0,b=1)
```

$\lambda=10$:

```{r}
par(mfrow=c(1,2))
k <- seq(0,1e+28,1e+25)
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=10),lambda=10),col="red")
abline(a=0,b=1)

k <- seq(0,1e+29,1e+26)
plot(k,inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=10,log=TRUE)),lambda=10),col="red")
abline(a=0,b=1)
```


$\lambda=100$:

```{r}
k <- seq(0,1000,1)
par(mfrow=c(1,2))
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=100),lambda=100),col="red")
abline(a=0,b=1)
plot(k,inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=100,log=TRUE)),lambda=100),col="red")
abline(a=0,b=1)
```
```{r}
k <- seq(0,10,0.1)
par(mfrow=c(1,2))
plot(k,inla.pc.qvminf(inla.pc.pvminf(k,lambda=100),lambda=100),col="red")
abline(a=0,b=1)
plot(k,inla.pc.qvminf(exp(inla.pc.pvminf(k,lambda=100,log=TRUE)),lambda=100),col="red")
abline(a=0,b=1)
```


reason for the issues of without exp(log()) when $\kappa$ is small:

```{r}
k <- seq(3,5,0.02)
inla.pc.pvminf(k,lambda=100)
```

```{r}
print("inla.pc.qvminf(1.110223e-16,lambda=100) returns:")
inla.pc.qvminf(1.110223e-16,lambda=100)
print("But:")
print("inla.pc.pvminf(4.039938,lambda=100)=")
inla.pc.pvminf(4.039938,lambda=100)
print("inla.pc.pvminf(4,lambda=100)=")
inla.pc.pvminf(4,lambda=100)
print("inla.pc.pvminf(3.91,lambda=100)=")
inla.pc.pvminf(3.91,lambda=100)
print("And, inla.pc.pvminf(3.9,lambda=100)=")
inla.pc.pvminf(3.9,lambda=100)
print("The value is too small that R regards it as 0.")
```