\documentclass[a4paper,11pt]{article}
\usepackage[scale={0.8,0.9},centering,includeheadfoot]{geometry}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{verbatim}

\begin{document}
\section*{New $0$inflated models: Poisson \& Binomial}

\subsection*{Parametrisation}

This is a new implementation (Nov'22) of zero-inflated Poisson and
Binomial likelihood, where we will allow for a linear predictor in
both the zero-inflattion and in the mean, but one of them needs to
consists of fixed effects only. This means the setup will be
somewhat different that for other likelihood models. 

\subsubsection*{Details}

The zero-inflated likelihood $f_0(y|\ldots)$ is defined as
\begin{displaymath}
    f_0(y | \eta_1, \eta_2) = p(\eta_1) 1_{[y=0]} +
    (1-p(\eta_1)) f(y | \eta_2)
\end{displaymath}
where $f(y|\ldots)$ is either Poisson or Binomial. We allow for two
linear predictors in the model, but one needs to be ``simple'' (i.e.\
only consists of fixed effects). The other is general and defined via
the formula. Normally, the zero-inflation probability is simpler (\texttt{family="0..."})
\begin{displaymath}
    \eta_1 = \text{simple} \qquad \eta_2 = \text{formula}
\end{displaymath}
but they can also be swapped (\texttt{family="0...S"})
\begin{displaymath}
    \eta_1 = \text{formula} \qquad \eta_2 = \text{simple}
\end{displaymath}

\subsection*{Link-function}

This is similar to Poisson and Binomial.

The link-function for the 'simple'-model must be given by argument
\texttt{link.simple} in the \texttt{control.family}-argument. Only
link-models without covariates/parameters are currently available. 
The examples later on shows how this is done.

\subsection*{Hyperparameters}

All parameters in the simple model are treated as hyperparameters.
The $j$'th element of $\eta_1$ is
\begin{displaymath}
    (\eta_1)_j = \sum_{i=1}^{m} \beta_i x_{ij}
\end{displaymath}
for covariates $x_1, \ldots, $, where $m$ is maximum 10. An intercept
in this model have to be defined manually by adding a constant
covariate vector.

\subsection*{New $1$inflated models for Poisson}

There are two similar models for the Poisson based on  the ``conditioned to be
positive'' Poisson $f(y | y > 0)$. These ones are named
\texttt{"1poisson"} and \texttt{"1poissonS"} and are 1-inflated.


\subsection*{Specification}

\begin{itemize}
\item \texttt{family="0poisson"}
\item \texttt{family="0poissonS"}
\item \texttt{family="0binomial"}
\item \texttt{family="0binomialS"}
\item \texttt{family="1poisson"}
\item \texttt{family="1poissonS"}
\item Required arguments: As for the Poisson and Binomial (but how
    these arguments are given, will differ). Optional argument
    \texttt{link.simple}.
\end{itemize}

\clearpage
\subsubsection*{Hyperparameter spesification and default values}
\paragraph{0poisson}
{\small \input{../hyper/likelihood/0poisson.tex}}
\clearpage
\paragraph{0poissonS}
{\small \input{../hyper/likelihood/0poissonS.tex}}
\clearpage
\paragraph{0binomial}
{\small \input{../hyper/likelihood/0binomial.tex}}
\clearpage
\paragraph{0poisson}
{\small \input{../hyper/likelihood/0binomialS.tex}}
\clearpage
\paragraph{1poisson}
{\small \input{../hyper/likelihood/1poisson.tex}}
\clearpage
\paragraph{1poissonS}
{\small \input{../hyper/likelihood/1poissonS.tex}}


\clearpage
\subsection*{Example: 0Poisson}
{\small \verbatiminput{example-0inflated-poisson.R}}

\clearpage
\subsection*{Example: Binomial}
{\small \verbatiminput{example-0inflated-binomial.R}}

\clearpage
\subsection*{Example: 1Poisson}
{\small \verbatiminput{example-1inflated-poisson.R}}


\end{document}

% LocalWords:  np Hyperparameters Ntrials gaussian hyperparameter

%%% Local Variables: 
%%% TeX-master: t
%%% End: 

