\documentclass[a4paper,11pt]{article}
\usepackage[scale={0.8,0.9},centering,includeheadfoot]{geometry}
\usepackage{amstext}
\usepackage{verbatim}

\begin{document}
\section*{The \texttt{gammacountmean} likelihood}

\subsection*{Parametrisation}

The \texttt{gammacount\_mu} distribution is a discrete probability
distribution on $0, 1, 2, 3, \ldots$, derived from a renewal process
with Gamma-distributed inter-arrival times. The probability mass
function is
\begin{displaymath}
    \text{Prob}(y) = G(T; y\alpha, \beta) - G(T; (y+1)\alpha, \beta)
\end{displaymath}
where $G(\cdot; a, b)$ is the cummulative distribution function of the
Gamma distribution, with shape $a>0$ and rate $b>0$,
\begin{displaymath}
    G(T; a, b) = \frac{b^a}{\Gamma(a)} \int_0^{T} u^{a-1}e^{-bu} du.
\end{displaymath}
This model exhibits:
\begin{itemize}
\item underdispersion if $\alpha > 1$ ($\mathrm{Var}(Y) < \mu$),
\item equidispersion if $\alpha = 1$ (Poisson limit),
\item overdispersion if $0 < \alpha < 1$ ($\mathrm{Var}(Y) > \mu$).
\end{itemize}

\subsection*{Link-function}

The linear predictor $\eta$ is linked to the event rate
$\lambda = {\texttt E}[Y]/T$ via
\begin{displaymath}
    \log(\lambda) = \eta,
\end{displaymath}
and the Gamma rate parameter $\beta$ is related to $\lambda$ through
the mean-parametrized renewal link
\begin{displaymath}
    \beta = \alpha \lambda + \frac{1}{2T}.
\end{displaymath}
This formulation accommodates unequal observation windows $T_i$ and
provides direct interpretation of $\exp(\eta_j)$ as the multiplicative
change in event rate per unit time.

\subsection*{Hyperparameter}

The dispersion parameter $\alpha > 0$ is modeled as
\begin{displaymath}
    \alpha = \exp(\theta),
\end{displaymath}
with a prior defined on the log-scale parameter $\theta$.

\subsection*{Specification}

\begin{itemize}
\item \texttt{family = "gammacountmean"}
\item Required arguments: response $y$ and exposure time $T$, provided as\\
    \texttt{inla.mdata(y, T)} (see example).
\end{itemize}

\subsubsection*{Hyperparameter specification and default values}
\input{../hyper/likelihood/gammacountmean.tex}

\subsection*{Example}

In the following example we estimate the parameters in a simulated
dataset with varying exposure times.
\verbatiminput{example-gammacountmean.R}

\subsection*{Notes}

\begin{itemize}
\item This model extends the classical Gamma-count framework
    (Winkelmann, 1995; Zeviani et al., 2014) by directly regressing on
    the event rate $\lambda$, unlike the original \texttt{gammacount}
    family in \texttt{R-INLA} (Nadifar et al., 2023), which assumes
    fixed $T=1$ and models waiting times.
\item The approximation
    ${\texttt E}[Y] \approx \beta T / \alpha - 1/(2\alpha)$ is
    asymptotically valid for moderate-to-large $T$ (Smith, 1958; Cox,
    1962).
\end{itemize}

\section*{References}
\begin{itemize}
\item  Cox, D. R. (1962). \textit{Renewal Theory}. Methuen.
\item  Nadifar, M., Baghishani, H., Fallah, A. (2023). A flexible
    generalized Poisson likelihood for spatial counts constructed by
    renewal theory, motivated by groundwater quality assessment.
    \textit{Journal of Agricultural, Biological, and Environmental
        Statistics}, 28(4), 726-748.
\item  Smith, W. L. (1958). Renewal theory and its ramifications.
    \textit{Journal of the Royal Statistical Society: Series B},
    20(2), 243–302.
\item  Winkelmann, R. (1995). Duration dependence and dispersion in
    count data models. \textit{Journal of Business \& Economic
        Statistics}, 13(4), 467–474.
\item  Zeviani, W. M., et al. (2014). The Gamma-count distribution
    in the analysis of experimental underdispersed data.
    \textit{Journal of Applied Statistics}, 41(7), 1513–1527.
\end{itemize}
\end{document}
